opponent modelling

### 2020-01-13 Report/draft_report_v1.Rmd
added MBM and stayes_EWA

As a measure for learning transfer, we focus on participants’ scores in the earliest rounds, excluding the very first round for which the agent is programmed to play randomly as it has no data on prior rounds on which to build its response. A group of players with no experience of the game are expected to have scores not significantly different from 0. Any significantly positive group average scores would therefore reflect prior learning from past experiences. In figure XXX, the average score across participants by game for rounds 2-6 are plotted. Scores are also averaged across levels of condition. We test the average scores for each game against a hypothesised value of 0 for a non-experienced player using parametric one sample t-tests. As expected for the initial RPS game, the average score is not significantly different from 0 ( t(148.85) = 1.04 , p = 0.89). In FWG, the score is significantly higher than 0 ( t(148.85) = 4.58 , p < 0.0001). Unexpectedly, this is also the case for NUMBERS ( t(148.85) = 3.00, p = 0.0092).

>>>>

As a measure for learning transfer, we focus on participants’ scores in rounds 2-6, excluding the very first round for which the agent is programmed to play randomly as it has no data on prior rounds on which to build its response. A group of players with no experience of the game are expected to have scores not significantly different from 0. Any significantly positive group average scores would therefore reflect prior learning from past experiences. In figure XXX, the average score across participants by game for rounds 2-6 are plotted. Scores are also averaged across levels of condition. We test the average scores for each game against a hypothesised value of 0 for a non-experienced player using parametric one sample t-tests. As expected for the initial RPS game, the average score is not significantly different from 0 ( t(148.85) = 1.04 , p = 0.89). In FWG, the score is significantly higher than 0 ( t(148.85) = 4.58 , p < 0.0001). Unexpectedly, this is also the case for NUMBERS ( t(148.85) = 3.00, p = 0.0092).

### 2020-01-13 comp_model_1019.Rmd
added MBM and stayes_EWA

# Fitting a bayesiam model WITH TRANSFER of opp strat distribution between games 
```{r}

@@

@@

## Putting results together 
```{r}
load("TR_bayes_modelling.RData")
load("NT_bayes_modelling.RData")
load("QL_modelling.RData")
load("QL_states_modelling.RData")
load("EWA_modelling.RData")
load("EWA_self_modelling.RData")

@@

                         # Self-Tuning EWA (only 1 parameter)
                          # Parametric EWA BIC
                         "EWA_self_2LL" = EWA_self_modelling[[id]]$optim$bestval,
                         "EWA_self_BIC" = EWA_self_modelling[[id]]$optim$bestval + 1*log(150),
                         #Lambda is a parameter of the softmax choice function (inverse Temperature)
                         "EWA_self_Lambda" = EWA_self_modelling[[id]]$optim$bestmem[1]

@@

## Building table of best model per participant
```{r}
Table_results <- table(All_results[, "condition"],c("random","Bayes Tr","Bayes No Tr","QL", "QL_states","EWA","S_EWA")[apply(All_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_BIC","QL_states_BIC","EWA_BIC","EWA_self_BIC")],1,which.min)])

@@

```
##### Looks like most participants behavior can be explained by Bayes with or without transfer. Most others are consistent with random behavior, and only 1 with QL and 1 with S_EWA.

@@

### Model based model: compute transition probabilities to new states. Weigh the value of an action by the conditional probabilty of ending up in a future state.... once you take an action in current round, you can only end up in 3 states, with probabilities p, q and 1-p-q. You are learning these probabilities by updating priors (initial prior is dirichlet). 
```{r}
MBM <- function(par,data,return_value,gamma){
  # Par[1] -> beta= inverse temperature parameter in softmax choice function
  # Par[2] -> lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # return_value=c("-2loglik","likelihood_by_trial")
  # Gamma is discount factor for future rewards 
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  alpha <- par[2]

  # Initiate probability vectors

>>>>

# Fitting a bayesian model WITH TRANSFER of opp strat distribution between games 
```{r}

@@

    # Update state
    curr_state <- new_state 
    #}
  }
  if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret) || is.nan(ret)) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)

@@

##  Self-Tuning EWA with prior round play as state (Camerer & Ho 2007) 
```{r}

EWA_states <- function(par,data,return_value){

  A_RPS = matrix(-0.5,9,3)
  dimnames(A_RPS) = list(states_RPS, c("R", "P", "S"))

  A_FWG = matrix(-0.5,9,3)
  dimnames(A_FWG) = list(states_FWG, c("F", "W", "G"))

  A_NUM = matrix(-0.5,25,5)
  dimnames(A_NUM) = list(states_NUM, c("1", "2", "3", "4", "5"))

  # # Define reward matrices from the perspective of the row player (human in our case)
  reward_RPS <- t(matrix(c(0,-1,1,1,0,-1,-1,1,0),nrow=3,ncol=3))
  dimnames(reward_RPS) = list(c("R", "P", "S"), c("R", "P", "S"))

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      Att <- switch(as.character(t_game),rps=A_RPS,fwg = A_FWG, numbers = A_NUM)
      state_vec <- switch(as.character(t_game), rps=states_RPS, fwg = states_FWG, numbers = states_NUM)
      reward <- switch(as.character(t_game), rps=reward_RPS, fwg = reward_FWG, numbers = reward_NUM)
      game_data <- subset(data, game == t_game)
      # first round is uniform prediction
      lik_hum[t] <- 1/nopts
      N <- 1.0

      # Randomly select prev_state and actions for first round
      curr_state <- sample(state_vec, size = 1)
      h_act_prev <- sample(colnames(Att), size =1)
      a_act_prev <- sample(colnames(Att), size =1)
      #reward <- 0

      indx <- data[t,"round"]
      state_indx <- match(curr_state,state_vec)
      #cat("this is the current state index", state_indx, "\n")

      # Get reward and past human action
      h_act_prev <- as.character(game_data[indx-1,"h_action"])
      a_act_prev <- as.character(game_data[indx-1,"a_action"])
      new_state <- paste0(h_act_prev,a_act_prev)

      # Estimate phi(t)
      R_t <- as.numeric(colnames(Att) == a_act_prev)
      #cat("This is R(t)",R_t,"\n")

      H_t <- (H_t*(indx-2) + R_t)/(indx-1)
      #cat("this is H(t)",H_t,"\n")

      Phi_t <- 1 - 0.5*sum((H_t-R_t)^2)
      #cat("This is Phit(t)",Phi_t,"\t")

      for (i in 1:nopts) {
        action <- as.character(colnames(Att)[i])

        #cat("this is current indexed action",action,"\n")
        #cat("this is previous human action",h_act_prev,"\n")

        # Attraction vector update rule
        Att[state_indx,i] <- (Phi_t*N*Att[state_indx,i]  + ( delta_t[[i]] + (1- delta_t[[i]])*(action == h_act_prev))*as.numeric(reward[action,a_act_prev])) / (Phi_t*N + 1)
      }

      #Update the value of N 
      N <- Phi_t*N + 1

      # Assume human chooses action probabilistically using softmax on Attraction values
      probs <- exp(lambda*Att[new_state,])/sum(exp(lambda*Att[new_state,]))

      # Get actual human action and compute likelihood
      h_act <- as.character(game_data[indx,"h_action"])
      act_index <- match(h_act, colnames(Att))
      lik_hum[t] <- probs[[act_index]]

      curr_state <- new_state

      # if (indx == 50) {
      #   cat("this is the attraction matrix at last round of", "\n")
      #   print(Att)
      # }
    }
  }

```{r}
# data = subset(dat,human_id == "38VxtUSv_h6RR5-tAAA2")
# EWA_states(0.1, data, "-2loglik")
# optim(10.0,fn=EWA_states,gr = NULL, data=data,"-2loglik", lower = 0.0, upper = 200, method="L-BFGS-B")

EWA_states_modelling <- list()
for(id in unique(dat$human_id)) {
  EWA_states_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
#   EWA_states_modelling[[id]] <- optim(10.0,fn=EWA_states,gr = NULL, data=tdat,"-2loglik", lower = 0.0, upper = 200, method="L-BFGS-B")
# }
  EWA_states_modelling[[id]] <- DEoptim(fn=EWA_states, lower = 0.0, upper = 100.0, data=tdat, "-2loglik", control=list(trace = FALSE,parallelType=1))
}

save(EWA_states_modelling,file="EWA_states_modelling.RData")

### Model based model: learn transition probabilities to new states. Weigh the value of an action by the conditional probability of ending up in a future state times the value of the new state.... once you take an action in current round, you can only end up in 3 states, with probabilities p, q and 1-p-q. You are learning these probabilities by updating priors (initial prior is dirichlet). 
```{r}
MBM <- function(par,data,return_value,gamma){
  # Par[1] -> beta= inverse temperature parameter in softmax choice function
  # Par[2] -> lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # return_value=c("-2loglik","likelihood_by_trial")
  # Gamma is discount factor for future rewards 
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  alpha <- par[2]

  # Transition Matrices 
  Transit_RPS = matrix(0,9,9)
  dimnames(Transit_RPS) = list(states_RPS, states_RPS)
  Transit_FWG = matrix(0,9,9)
  dimnames(Transit_FWG) = list(states_FWG, states_FWG)
  Transit_NUM = matrix(0,25,25)
  dimnames(Transit_NUM) = list(states_NUM, states_NUM)

    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, numbers = Q_vals_NUM)
      state_vec <- switch(as.character(t_game),rps=states_RPS,fwg = states_FWG, numbers = states_NUM)
      Transit <- switch(as.character(t_game),rps=Transit_RPS,fwg = Transit_FWG, numbers = Transit_NUM)
      lik_hum[t] <- 1/nopts
      # Randomly select prev_state and actions for first round
      curr_state <- sample(state_vec, size = 1)
      h_act <- sample(colnames(Q_vals), size =1)
      ai_act <- sample(colnames(Q_vals), size =1)
      reward <- 0

    # Update Transit matrix ....first create hot vector = 1 if new state, 0 otherwise 
    hot_vector <- state_vec == new_state

    # Transit probs update: TD learning on transition matrix, learning rate fixed at 0.1
    Transit[curr_state,] <- Transit[curr_state,] + 0.1* (hot_vector - Transit[curr_state,])
    #cat(Transit[curr_state,], "\n")

    # Update Q-values as current reward + a weighted (by trasnsit probs) average of future Q-values.
    Q_Row_max <- apply(Q_vals, 1, max, na.rm = TRUE)
    #cat(Transit[curr_state,] * Q_Row_max, "\n")
    Q_vals[curr_state, h_act] <- reward +  gamma*( Transit[curr_state,] %*% Q_Row_max  )

```{r}
MBM_modelling <- list()
for(id in unique(dat$human_id)) {
  MBM_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  # MBM_modelling[[id]] <- optim(c(1,0.1),fn=QMBM,gr = NULL, data=tdat,"-2loglik", gamma =0 , lower = c(0,0), upper = c(10,0.99), method="L-BFGS-B")

   MBM_modelling[[id]] <- DEoptim(fn=MBM,lower = c(0,0), upper = c(10,1), data=tdat,"-2loglik", gamma = 0.9, control=list(trace = FALSE,parallelType=1))
}

save(MBM_modelling, file="MBM_modelling.RData")

## Putting results together 
```{r}
load("TR_bayes_modelling.RData")
load("NT_bayes_modelling.RData")
load("QL_modelling.RData")
load("QL_states_modelling.RData")
load("EWA_modelling.RData")
load("EWA_self_modelling.RData")
load("MBM_modelling.RData")
load("EWA_states_modelling.RData")

@@

                         # Self-Tuning EWA (only 1 parameter)
                          # Parametric EWA BIC
                         "EWA_self_2LL" = EWA_self_modelling[[id]]$optim$bestval,
                         "EWA_self_BIC" = EWA_self_modelling[[id]]$optim$bestval + 1*log(150),
                         #Lambda is a parameter of the softmax choice function (inverse Temperature)
                         "EWA_self_Lambda" = EWA_self_modelling[[id]]$optim$bestmem[1],

                         # EWA with states

                         "EWA_states_BIC" = EWA_states_modelling[[id]]$optim$bestval + 1*log(150),
                         "EWA_states_Lambda" = EWA_states_modelling[[id]]$optim$bestmem[1],

                         # MBM 
                          # MBM BIC
                         "MBM_BIC" = MBM_modelling[[id]]$optim$bestval + 2*log(150),
                          # BEta parameter inverse temp in softmax 
                         "MBM_beta" = MBM_modelling[[id]]$optim$bestmem[1],
                         "MBM_alpha" = MBM_modelling[[id]]$optim$bestmem[2]

@@

## Building table of best model per participant
```{r}
Table_results <- table(All_results[, "condition"],c("random","Bayes Tr","Bayes No Tr","QL", "QL_states","EWA","S_EWA","EWA_states", "MBM")[apply(All_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_BIC","QL_states_BIC","EWA_BIC","EWA_self_BIC","EWA_states_BIC","MBM_BIC")],1,which.min)])

@@

```
##### Looks like most participants behavior can be explained by QL_states. Bayesian models best explain a few more behaviors, while the others are consistent with random behavior, and only 1 with QL and 1 with S_EWA.

@@
